Human Factors Influencing Visual Statistical Inference

### 2013-06-06 sociology_paper.tex
start of intro revised

Statistical graphics can be used to test the significance of findings during the exploratory data analysis. The method is now termed as visual statistical inference which is discussed in the following section.

\subsection{Visual Statistical Inference} The fundamental concept of visual statistical inference is introduced by \citet{buja:2009}. In their paper, they proposed lineup protocol to discover significant pattern in the data. Later these concepts are further developed by defining necessary terminologies and validating using a head to head comparison with conventional inference by \citet{majumder:2013}. The power of the visual test is obtained and it is observed that the power is as good as that of conventional test while in some scenarios, visual power exceeds that of conventional tests.

The test statistic used in visual inference is a plot of the observed data. This plot, called actual data plot, is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the actual data plot in the lineup, the hypothesis is rejected. Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which of these towers is the most extreme? Which side is higher? In an experiment 12 of 72 observers picked plot \# $3^2+4$ as the most extreme plot from the lineup, based on the highest amount of asymmetry (36\%), a trend (26\%) or `outliers' (16\%). 
This gives us a visual $p$-value of 0.00023, indicating that we have enough evidence to reject the null hypothesis (in another four repetitions we found similar results, giving an overall $p$-value of less that $10^{-17}$ based on 62 data picks out of 341 evaluation). 
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots: in this example, we are investigating, whether the results from the 2012 US presidential election is consistent with polling results. As a null hypothesis we  assume that  the election results are consistent with polling results. The polling results give us a null model to generate data. Each poll gives us a margin between candidates. This is what we use as the mean of a normal distribution. In the absence of information of error for a poll, we will assume a standard deviation of 2.5 percentage points for all of the polls. While this is a severe simplification, the resulting margin of error of $\pm 5$ percentage points is realistic, and could be easily replaced if additional information on a poll became available. A null data set is then generate as a set of draws from normal distributions based on the latest poll results. Each panel in figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes}: each state is represented by a rectangle. The width of a rectangle is determined by the difference between (polling) results, its height is given by the number of electoral votes, and fill color shows party affiliation (blue for Democrat, and red for Republican). Rectangles are placed on top of each other, separated by affiliation and sorted by margin between polls. A party needs 270 electoral votes or more to win the presidential election.

>>>>

The lineup protocol introduced in \citet{buja:2009} can be used to test the significance of findings during the exploratory data analysis. The methodology is a part of what is called visual statistical inference.  These concepts have been developed further by \citet{majumder:2013} who refined the terminology and validated the lineup protocol with a head to head comparison with conventional inference. One of the major contributions of  \citet{majumder:2013}  is to define the power of the visual test and how to estimate the power for a particular lineup. It was observed that the power can be as good or better than that of a conventional test in some scenarios.

In visual inference, the test statistic is a plot of the observed data. To create a lineup, this plot, called actual data plot, is placed on a layout of null plots. The null plots  are generated from the model specified by a null hypothesis, essentially describing what the plot might look like if the data had no structure. An observer is asked to evaluate the lineup. If the  actual data plot is detected by the observer, the null hypothesis is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness.

\subsection{Visual inference example}
 Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which of these towers is the most extreme? Which side is higher? In an experiment 12 of 72 observers picked plot \# $3^2+4$ as the most extreme plot from the lineup, based on the highest amount of asymmetry (36\%), a trend (26\%) or `outliers' (16\%). 
This gives us a visual $p$-value of 0.00023, indicating that we have enough evidence to reject the null hypothesis (in another four repetitions we found similar results, giving an overall $p$-value of less that $10^{-17}$ based on 62 data picks out of 341 evaluation). 
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots: in this example, we are investigating, whether the results from the 2012 US presidential election is consistent with polling results. As a null hypothesis we  assume that  the election results are consistent with polling results. The polling results give us a null model to generate data. Each poll gives us a margin between candidates. This is what we use as the mean of a normal distribution. In the absence of information of error for a poll, we will assume a standard deviation of 2.5 percentage points for all of the polls. While this is a severe simplification, the resulting margin of error of $\pm 5$ percentage points is realistic, and could be easily replaced if additional information on a poll became available. A null data set is then generate as a set of draws from normal distributions based on the latest poll results. Each panel in figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes}: each state is represented by a rectangle. The width of a rectangle is determined by the difference between (polling) results, its height is given by the number of electoral votes, and fill color shows party affiliation (blue for Democrat, and red for Republican). Rectangles are placed on top of each other, separated by affiliation and sorted by margin between polls. A party needs 270 electoral votes or more to win the presidential election.
